Real-time Acoustic Ray Range Estimation for Underwater Vehicle Navigation

### 2021-08-21 doc/main.tex
updates, need to finish intro, proofread

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-30.pdf}
  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m. The beacons are highlighted in color/marker coding in figure \ref{fig:overview}. The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative receiver colors over a total ray fan in gray.}
  \label{fig:raytrace-zs30}
  \end{figure}

@@

As shown in the eigenray traces, an acoustic arrival does not always take the direct path from source to the receiver, and the path it takes is extremely dependent on the sound speed profile.
Often, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
The suggested algorithm, the nearest bounce criteria (NBC), is only a slight modification from the minimal bounce criteria, and includes multipath as a new dimension of information with negligible additional computation.

@@

\begin{table}[h!]
\renewcommand{\arraystretch}{1.3}
\centering
\begin{tabular}{r|c|c|c|c|c}
SSP & Count & ~Mean [m] & ~Median [m] & ~STD [m] & ~Max [m] \\ \hline
Baseline & 243 & 11.38 & 11.96 & 4.23 & 23.95 \\ 
Chosen weights & 568 & 11.36 & 11.40 & 8.12 & 26.0 \\
\end{tabular}
\caption[Range estimation error for \textit{in situ} events]{The range estimation error for \textit{in situ} events. The error here is presented as the absolute, not directional, error. Almost all directional errors, however, are positive.}
\label{tab:rangeErrorInSitu}
\end{table}

@@

A post-processing analysis that replicated the real-time pipeline was run to overcome the unequal distribution of communication events with respect to depth, range, and sound speed status.
There are 1242 total beacon to beacon events in the modem experiment that would have been accepted by the ICNN for real-time ranging and navigation had the vehicle behavior been running on all modems.
The value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
Naturally, because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
The analysis here therefore seeds ``omniscient'' knowledge of the range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN.

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{figs/SLIDES-gvel-comparison-wdata-30.pdf}
\caption{\label{fig:gvel30}{A comparison of group velocity predictions for all beacon to beacon events in post-processing with a source depth of 30 m. The sound speed source is indicated by color. The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{figs/compare-methods-postv1v2.pdf}
\caption{\label{fig:compareV1V2}{Caption here.}}
\end{figure}

If there was an assumed deterministic group velocity, the range error would grow with OWTT proportionally to the difference between the estimated and actual value.
Thus, an indication that the algorithm is adapting group velocity correctly is that range error does not strictly increase with OWTT.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{figs/range-error-owtt-newalgorithm.pdf}
\caption{\label{fig:gvel30}{The post-processed range error for source depths of 20, 30, and 90 m, and receiver depths of 30 and 90 m. The dashed gray line shows no error. The shaded region connects the range performance across all events.}
\end{figure}

\FloatBarrier
\subsection{GPS sensor drift}

@@

Range estimation is an essential step of acoustic localization and navigation.
This work solely focuses on mitigating range estimation error via a stochastic prediction of the group velocity, knowing that this can have an disproportionate benefit on minimizing trilateration error.
We hypothesize and validate that the embedded stochastic prediction of a single group velocity is a smoothly varying function of range, source and receiver depth pairings, and multipath structure.
We employ a GPS-tracked experiment to have a ground truth comparison for real-time localization algorithms.

The real-time system achieves GPS-like navigation without taking into account multipath structure; the ranging error improves by an order of magnitude with the suggested multipath adaptability, minimizing range error to single meters. 
Post-processing analysis shows that this method of ranging is actually sensitive to GPS drift.

This work, to our knowledge, is the first successful field deployment of a real-time, ray-based acoustic ranging system for underwater navigation.
The methods here demonstrate remarkable ranging accuracy and precision for the Beaufort Lens, an acoustically complex environment.
The \textit{in situ} prediction via BELLHOP balances the tension in simulation fidelity versus capturing the overarching ensemble physics to approximate a local group velocity.

Furthermore, the improved nearest bounce criteria, validated in a high fidelity post-processing pipeline, produces range estimates that rival GPS accuracy and outclass GPS precision, especially at polar latitudes.
It transforms multipath, widely considered as an obstacle for acoustic ranging, into a new information content to refine ranging accuracy.

The scalability of the methods provided here for long-range applications remain to be seen.
Given the NBC algorithm's improvement with increased multipath, its effectiveness is likely only challenged by the valid operational scales of a range independent propagation environment.
Even then, the dominant SSP amongst the variability may very well produce GPS-like accuracy.
The NBC algorithm can be easily be retooled to discriminate between surface and bottom bounces, and employ simple statistical learning or more advanced machine learning techniques to select an appropriate representative group velocity.
Some more advanced vehicle intelligence can toggle use the range uncertainty to toggle between a more conservative group velocity estimate and the NBC, which is intrinsically quite precise.

There are many avenues that this approach can be further refined and tested for field operations.
Amongst them is defining the uncertainty grid for BELLHOP via stochastic or data-driven measures such as the distance traveled by the AUV between ICNN updates or the magnitude of the position corrections by the ICNN.
Another is to entirely remove the seeded range and instead rely on the submerged asset's depth and recorded OWTT to find high probability fields in range, as briefly indicated in Chapter 6.

As the Arctic continues to become more accessible and navigable, safe navigation infrastructure and practices will be increasingly vital for industry, tourism, commercial, and military operations.
This work contributes a new paradigm to embed environmentally adaptive behavior into AUV navigation that has demonstrated success in an acoustically complex and operationally challenging environment.
Thus, we believe that this work enables more accurate range estimation, localization, and/or navigation for any field experiment given known source and receiver depths.

(NEED TO SKIM CONCLUSION FOR MORE MATERIAL)

@@

\clearpage
\FloatBarrier
\subsection*{Group velocity prediction by source depth}
\begin{figure}[h!]
\includegraphics[width=0.8\textwidth]{figs/SLIDES-gvel-comparison-wdata-20.pdf} \\
\vspace{2em}
\includegraphics[width=0.8\textwidth]{figs/SLIDES-gvel-comparison-wdata-90.pdf}
\caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing. The top panel shows a source depth of 20 m; the bottom panel shows a source depth of 90 m. The sound speed source is indicated by color. The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}
\end{figure}

>>>>

Autonomous underwater vehicles (AUVs) are an increasingly capable platform to explore and sample the ocean, particularly for remote and/or dangerous environments.
Lightweight, mobile, and inexpensive compared to crewed- and ship-based systems, these vehicles can enable a safer, data-forward style of field oceanography \citep{bellingham_robotics_2007,petillot_underwater_2019}.

But navigation uncertainty is a major challenge in considering AUVs as reliable and standard tools for oceanographic research.
While land and air-based robots utilize satellite-based global positioning system (GPS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GPS while underwater due to the rapid attenuation of electromagnetic waves \citep{preisig_acoustic_2007}.
A vehicle can stall on the surface to receive a GPS fix, but this foolproof method of re-positioning is inaccessible in an ice-covered environment or other GPS-denied situations.
A recent study tracking 22 floats under the ice in the Weddell Gyre near Antarctica showed that they were unable to surface for 8 months and position uncertainty grew to 116 $\pm$ 148 km in the lateral directions \citep{chamberlain_observing_2018}.

Much work has been done in underwater vehicle navigation, which relies on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and/or acoustic baseline navigation systems \citep{paull_auv_2014}.
In particular, the long baseline (LBL) navigation system is the most similar in scale and style to GPS, and most appropriate for mitigating drift in underwater vehicle localization without overburdening computation on the vehicle \citep{van_uffelen_global_2021}.

\subsection{Timing, Positioning, and Navigation}

Underwater acoustic navigation (and GPS) is predicated on timing and positioning.
Timing is the ability to maintain accurate and precise time, sometimes synchronized across multiple nodes.
The advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016} enabled ranging via one-way travel time (OWTT).

Positioning is the ability to accurately determine one's location and is generally done via trilateration.
Real-time navigation efforts assume a locally homogeneous sound speed profile (SSP) that enables a linear, deterministic scaling between travel time and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
Short range (less than ten kilometer) navigation efforts accurately rely on a nominal sound speed value and achieve minimal navigation error \cite{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
For mesoscale (tens to hundreds of kilometers) navigation or localization efforts, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed is used.

Post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), without computational or time constraints, can indulge more complex acoustic methods and/or ocean representations.
There are many different approaches, from accounting for a depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}, or isolating a specific ray path to find a representative group velocity \citep{mikhalevsky_deep_2020}.
The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.

\subsection{Underwater positioning and navigation in the Arctic}

However, the key assumptions for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash did not hold during a vehicle deployment in ICEX16, in the Beaufort Sea, in March 2016.

However, the new Arctic ... [Beaufort Lens]

The literature in underwater acoustic navigation and positioning is either real-time or physics-based.
AUV operations in the Beaufort Sea necessitate a real-time and physics-based method to overcome the unique challenges of total ice cover, multiple ice floes, and the Beaufort Lens.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-30.pdf}
  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m. The beacons are highlighted in color/marker coding in Fig. \ref{fig:overview}. The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative receiver colors over a total ray fan in gray.}
  \label{fig:raytrace-zs30}
  \end{figure}

@@

As shown in the eigenray traces of Fig. \ref{fig:raytrace-zs30}, an acoustic arrival does not always take the direct path from source to the receiver, and the path it takes is extremely dependent on the sound speed profile.
Often, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
The suggested algorithm, the nearest bounce criteria (NBC), is a slight modification from the minimal bounce criteria, and includes multipath as a new dimension of information with negligible additional computation.
This metric was introduced in post-processing.

@@

\begin{table}[h!]
\renewcommand{\arraystretch}{1.3}
\centering
\begin{tabular}{r|c|c|c|c|c}
SSP & Count & ~Mean [m] & ~Median [m] & ~STD [m] & ~Max [m] \\ \hline
Baseline & 243 & 11.38 & 11.96 & 4.23 & 23.95 \\ 
Chosen weights & 568 & 11.36 & 11.40 & 8.12 & 26.0 \\
\toprule
\end{tabular}
\caption[Range estimation error for \textit{in situ} events]{The range estimation error for \textit{in situ} events. The error here is presented as the absolute, not directional, error. Almost all directional errors, however, are positive.}
\label{tab:rangeErrorInSitu}
\end{table}

@@

There are 1242 total beacon to beacon events in the modem experiment that would have been accepted by the ICNN for real-time ranging and navigation had the vehicle behavior been running on all modems.
A post-processing analysis that replicated the real-time pipeline was run to overcome the unequal distribution of communication events with respect to depth, range, and sound speed status.

The value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
The analysis here therefore seeds realistic but ``omniscient'' knowledge of the range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN.

@@

The minimal and nearest bounce algorithm is applied using the three sound speed inputs shown in Fig.~\ref{fig:raytrace-zs30}; the resulting predicted group velocities for a source depth of 30 m are shown in Fig~\ref{fig:gvel30}.
The takeaways from this source depth are representative of those from 20 and 90 m as well (see Supplementary Fig.~\ref{fig:gvelMore}).

\begin{figure}[h!]
\includegraphics[width=\columnwidth]{figs/SLIDES-gvel-comparison-wdata-30.pdf}
\caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing with a source depth of 30 m, with group velocity on the y-axis and recorded travel time on the x-axis. The left panel is for a receiver depth of 30 m; the right panel for 90 m. The sound speed source is indicated by color. The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}
\label{fig:gvel30}
\end{figure}

Because direct path dominates the predictions from the MBC, group velocity estimates are greater than those from NBC.
The one instance this is not true is important to mention\textemdash for a travel time of 2.2 s, where the minimal multipath is dominated by a single bottom bounce, but the eigenray traces show two, three, and four bounces for HYCOM, baseline, and weighted SSPs, respectively (the red line to the furthest transmission in Fig. \ref{fig:raytrace-zs30}).
The nearest bounce method never classifies anything as direct path for the 30 m source, though BELLHOP is quite adept at finding those rays.

The goal of the group velocity estimation is to converge towards the naive group velocity estimation, i.e. the GPS-derived range divided by the recorded travel time.
For a 30 m receiver depth, the NBC shows more overlap with data-derived values as it classifies multipath more correctly.
For a 90 m receiver depth, the overlap is less accurate due to computational constraints.
Given a limited fan of rays entering the shadow zone, there is a less reliable simulated time arrival packet.

Overall, analysis shows that the higher multipath classification produces more accurate range estimates, likely driven by a tighter bundle of rays.
However, the data is too small to draw significant conclusions.
Discontinuities in multipath classification verify our hypothesis for its importance for a smoothly varying group velocity, as shown in the cluster where HYCOM jumps from one to two bounces amidst the baseline SSP seeing two bounces and the weighted SSP seeing three.

When using a linear scaling to convert travel time into range, any offset between the assumed sound speed and the horizontal group velocity produces unconstrained error with increasing receiver distance.
Figure \ref{fig:rangeError} shows the directional range error footprints for all three sound speed inputs with respect to OWTT, separated by source and receiver depth configurations.
The weighted SSP range error generally has the smallest and most zero-centered footprint.
The one case it does not is for the source-receiver pairings between 30 and 90 m in depth.
The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts propagating through the steep sound speed gradients of the lens, through a shadow zone. 
All other source depth pairings are significantly improved using the chosen weights compared to HYCOM or the baseline.
Most importantly, we see the consequences of the adaptive group velocity in that range error does not strictly increase with OWTT.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{figs/range-error-owtt-newalgorithm.pdf}
\caption{The post-processed range error for source depths of 20, 30, and 90 m, and receiver depths of 30 and 90 m. The dashed gray line shows no error. The shaded region connects the range performance across all events.}
\label{fig:rangeError}
\end{figure}

@@

The previous section displayed the psuedorange performance using the NBC.
Here, we explicitly compare the ranging accuracy between the nearest and minimal bounce criteria for all events in the post-processing pipeline.
It is important to note that because the post-processing pipeline is the same for both methods, that the efficacy of the NBC compared to the MBC is 100\%, as the former simplifies to the latter for direct path classifications.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{figs/compare-methods-postv1v2.pdf}
\caption{\label{fig:compareV1V2}{A comparison of range estimation error for all beacon to beacon events in post-processing. From left to right, the SSPs are the baseline, the chosen weights, and HYCOM. The colors indicate the source depth, darkening with depth, and the shapes indicate the multipath structure. Because this plot is square, the shaded region shows where the updated algorithm is less accurate than the \textit{in situ} algorithm.}}
\end{figure}

Figure \ref{fig:compareV1V2} is a crossmap for all three SSPs for both group velocity predictors, with the minimal bounce on the x-axis and the nearest bounce on the y-axis; the important range estimation metrics are summarized in table \ref{tab:rangeV1V2}.

The first thing to notice is that all the events are on the right hand side of the plots\textemdash the MBC has a systemic overestimation, which is reduced via NBC as events straddle the x-axis.
The improvement from MBC to NBC is most evident for the realistic sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
Table \ref{tab:rangeErrorPost} shows further statistics on the absolute range error by SSP and group velocity algorithm.
The order of magnitude improvement in the ducted SSPs demonstrate the effectiveness of the algorithm exploiting the multipath conditions.
The gridded approach and multipath classification build in the computational cushion for minimal error and minimal bias compared to direct eigenray assessment.

\begin{table}[h!]
\renewcommand{\arraystretch}{1.5}
\centering
\begin{tabular}{r|cc|cc|cc}\toprule
 & \multicolumn{2}{c|}{\textbf{Baseline} } & \multicolumn{2}{c|}{\textbf{Chosen Weights} } & \multicolumn{2}{c}{\textbf{HYCOM}} \\
 algorithm & \cellcolor[HTML]{EFEFEF}minimal & nearest & \cellcolor[HTML]{EFEFEF} minimal& nearest & \cellcolor[HTML]{EFEFEF}minimal & nearest \\ \hline
 minimum [m] 	& \cellcolor[HTML]{EFEFEF}0.01 & 0.00	& \cellcolor[HTML]{EFEFEF}0.00 	& 0.00 	& \cellcolor[HTML]{EFEFEF}0.11 & 0.01 \\
 25th \% [m]   & \cellcolor[HTML]{EFEFEF}4.96 & 0.99	& \cellcolor[HTML]{EFEFEF}6.26 	& 0.95 	& \cellcolor[HTML]{EFEFEF}3.30 & 2.25 \\   
 median [m]		& \cellcolor[HTML]{EFEFEF}10.30 & 2.27 	& \cellcolor[HTML]{EFEFEF}13.28 & 2.12 	& \cellcolor[HTML]{EFEFEF}6.41 & 4.61 \\
 75th \% [m]   & \cellcolor[HTML]{EFEFEF}15.81 & 5.51 	& \cellcolor[HTML]{EFEFEF}19.75 & 4.11 	& \cellcolor[HTML]{EFEFEF}10.92 & 7.46 \\
 maximum [m]    & \cellcolor[HTML]{EFEFEF}22.52 & 14.96 & \cellcolor[HTML]{EFEFEF}1491  & 20.21 & \cellcolor[HTML]{EFEFEF}19.55 & 15.81 \\
 \toprule
\end{tabular}
\caption[Comparison of post-processing range estimation algorithms across all events]{A comparison of range estimation metrics for each sound speed source and group velocity estimation algorithm for all 1283 beacon to beacon events via post-processing. The 0th (minimum), 25th, 50th (median), 75th, and 100th (maximum) percentiles are shown to the range resolution afforded by the WHOI Micro-Modem. There are a few outliers that drive the mean to be higher than the median.}
\label{tab:rangeErrorV1V2}
\end{table}

All plots show the same 45 degree banding of source depth and multipath structure, indicating systemic GPS noise.
Even with poorer accuracy from the GPS solution, the acoustic methods remain high precision.
Some bands are adjacent to others of the same styling, likely due to ice drift.

\FloatBarrier
\subsection{Further analysis of the anomalous beacon to beacon connection}

As shown in table \ref{tab:rangeErrorPost}, there is a striking maximum range error of 1491 m for weighted SSP in the minimal bounce criteria.
This section isolates the events that encompass that outlier for further examination, summarized in table \ref{tab:rangeErrorPostExplanation}.
There are 10 events from South transmitting at 30 m depth to North receiving at 30 m depth.
The OWTT spread is from 2.1958 to 2.1963 s; the naive group velocity is 1429.3 to 1430.1 m/s; and the GPS-tracked range is from 3138.54 m to 3140.87 m.
This example ends up being an excellent case study for how sound speed and multipath fidelity work in concert to minimize range error.

\begin{table}[h!]
\renewcommand{\arraystretch}{1.5}
\centering
\begin{tabular}{r|cc|cc|cc}\toprule
 & \multicolumn{2}{c|}{\textbf{Baseline} } & \multicolumn{2}{c|}{\textbf{Chosen Weights} } & \multicolumn{2}{c}{\textbf{HYCOM}} \\
 algorithm & \cellcolor[HTML]{EFEFEF}minimal & nearest & \cellcolor[HTML]{EFEFEF} minimal& nearest & \cellcolor[HTML]{EFEFEF}minimal & nearest \\ \hline
 \# bounces 				& \cellcolor[HTML]{EFEFEF}1 		& 3 		& \cellcolor[HTML]{EFEFEF}1 		& 4 		& \cellcolor[HTML]{EFEFEF}0 	 & 2 \\
 mean OWTT [s] 				& \cellcolor[HTML]{EFEFEF}2.1853 	& 2.1943 	& \cellcolor[HTML]{EFEFEF}4.1812 	& 2.1955 	& \cellcolor[HTML]{EFEFEF}2.1877 & 2.1902 \\
 mean group velocity [m/s] 	& \cellcolor[HTML]{EFEFEF}1436.7 	& 1430.8 	& \cellcolor[HTML]{EFEFEF}750.88 	& 1430.0 	& \cellcolor[HTML]{EFEFEF}1435.1 & 1433.4 \\
 range error [m] 			& \cellcolor[HTML]{EFEFEF}15.4 		& 2.39 	 	& \cellcolor[HTML]{EFEFEF}-1491 	& 0.77 		& \cellcolor[HTML]{EFEFEF}11.9 & 8.30 \\
 \toprule
\end{tabular}
\caption[Further look at algorithm differences for one beacon to beacon pairing]{A comparison of range estimation metrics for each sound speed source and group velocity estimation algorithm on all 10 communication events between North at 30 m \& South at 30 m.}
\label{tab:rangeErrorPostExplanation}
\end{table}

The large error is driven by the MBC unexpectedly defaulting to a bottom bounce with a much greater OWTT.
Thus the NBC ``correctly'' classifies the multipath as 4 bounces, reducing the range error from greater than a kilometer to less than a meter.
While there is no actual way of knowing if this is the correct multipath structure, the range error is remarkably small.
This pattern of not choosing the minimal observed bounce structure is consistent across all SSPs; the baseline goes from 1 to 3 bounces and HYCOM goes from 0 to 2 bounces.
Notably, the baseline and HYCOM range errors are never egregiously large, but are nonetheless improved with the NBC algorithm.
Thus, for acoustically complex environments, the NBC has a disproportionately positive impact as the estimated SSP approaches the desired SSP.

In comparison to the earlier eigenray traces, the results here distill the importance of the gridded approach to curate multipath propagation. 
The grid gives BELLHOP tolerance for finding solutions that otherwise may not be found in a center or single point solution; the limitations of numerical computation, especially in a real-time environment, are more adeptly addressed by baking in a small amount of uncertainty than trying to prescribe an exact solution.
This creates more data to sort through given that BELLHOP is a greedy estimator of eigenrays, but is computationally simpler than running a dense ray fan for precise eigenray finding.

The NBC algorithm, coupled with sound speed fidelity, work in concert to mitigate error.
Either one alone is not enough for the Beaufort Lens, as the environmental realism begins to match the acoustic conditions of intense multipath.

\FloatBarrier
\subsection{GPS sensor drift at polar latitudes}

@@

Range estimation is an essential step of acoustic localization and navigation.
Current approaches in real-time underwater acoustic navigation simplify the non-linear relationship between a sound speed profile and acoustic propagation with a deterministic sound speed.
Some post-processing approaches attempt to leverage the acoustic arrival structure via various ray methods, but often use a singular SSP for long-term deployments.
Thus, the conversion from travel time to range, particularly for real-time vehicle deployments, can be pre-conditioned for error growth as the OWTT/range increases.
For GPS-denied navigation, and hostile environments like the Arctic, tolerance for error is close to none.
This work addresses a critical need in acoustic navigation by retooling acoustic arrival methods generally deemed too complex or labor intensive for real-time, ray-based range estimation to achieve GPS-like positioning.

We hypothesize and validate that the embedded stochastic prediction of a single group velocity is a smoothly varying function of range, source and receiver depth pairings, and multipath structure.
We employ a GPS-tracked experiment to have a ground truth comparison for real-time localization algorithms.
The real-time system achieves GPS-like navigation for an AUV without taking into account multipath structure; the ranging error improves by an order of magnitude with the suggested multipath adaptability, minimizing range error to single meters.
Post-processing analysis shows that this method of ranging is sensitive to GPS drift itself; thus embedding a stochastic prediction of the horizontal group velocity has an outsized benefit to minimizing trilateration error.

There are many avenues that this approach can be further refined and tested for field operations.
Amongst them is defining the uncertainty grid for BELLHOP via stochastic or data-driven measures such as the distance traveled by the AUV between ICNN updates or the magnitude of the position corrections by the ICNN.
Another is to entirely remove the seeded range and instead rely on the submerged asset's depth and recorded OWTT to find high probability fields in range.

This work, to our knowledge, is the first successful field deployment of a real-time, model-based acoustic ranging system for underwater navigation.
The methods exploit the upward refracting nature and the total ice cover of the Arctic environment to achieve remarkable ranging accuracy and precision.
It transforms multipath, widely considered as an obstacle for acoustic ranging, into a new information content to refine ranging accuracy.
We believe that this work enables more accurate range estimation, localization, and/or navigation for any field experiment given known source and receiver depths.

\subsection{Broader relevance \& future work}

Performance in other acoustic environments will require introducing a different thresholded metric to sort time of arrivals.
Given the NBC algorithm's improvement with increased multipath, its effectiveness is likely only challenged by the valid operational scales of a range independent propagation environment.
For mesoscale operations, like that of many gliders, the group velocity criteria may need to be modified to better account for variability driven by range dependent propagation through internal waves, eddies, or even bathymetric changes like underwater volcanoes.
BELLHOP simulations provide other relevant eigenray information, like time and angle of arrival, that is ripe for statistical and machine learning methods to classify a representative group velocity.
A bespoke and fast ray tracing method, which was used during ICEX20 to evaluate the acoustic fit of sound speed profiles, can easily report back the number of turning points instead of the number of bounces for multipath classification.

This approach will start to break down in extremely dynamic environments.
Fast moving fronts, as seen in estuaries like the Connecticut River into the Long Island Sound, present an entirely new set of challenges not seen by internal waves or eddies.
Buoyancy fluctuations in these regions threaten even simple AUV tasks like following a trackline.
Acoustic communications are further complicated given a shallow environment with significant scattering \citep{lavery_measurements_2010,ross_acoustic_2012,lavery_broadband_2013}, where fast timefront modeling may only be coherent for trivial probabilities of the ocean state.
Realistic \textit{in situ} considerations of the acoustic environment may not be possible in such environments without complete through-the-sensor integration of echosounder data and/or a hyper-realistic onboard ocean model.

Many approaches to underwater navigation combine it with acoustic tomography, i.e., a joint estimation of both source and receiver locations and the ocean volume between them.
There has been considerable success at this effort in post-processing methods, which utilize intensive\textemdash and due to the non-linearity of sound propagation, often brute force\textemdash computational methods. 
For vehicle operations, fast tomography is the ideal implementation, in that one can fully consider how sound speed structure, horizontally and vertically, influences sound propagation.
AUVs can serve as moving sources to better image the ocean volume \citep{deffenbaugh_optimal_1997,elisseeff_ocean_2002}, where mobile tomography and navigation converge on the same set of component technologies: position estimation, sound speed parameterization estimation, ray path identification, and vehicle path optimization.

But there are overwhelming challenges, operationally and computationally, for fast, mobile tomography to become a realistic endeavor.
Addressing the spatial and temporal scales of what can be solved deterministically and what must be solved stochastically imposes a resolution constraint on the utility of gridded models that already dwarfs that of AUV operations.
Given that AUV operations are often on small spatial and temporal scales, the added benefit of a gridded model is quite small, and in cases like the Arctic, may actually mischaracterize the ocean volume.

For gliders, with longer and larger operational scales, an ocean model may provide useful information.
Currently gliders are low power and do not have the storage or computational power to run a full-scale, realistic ocean model.
Of course, all of this is assuming that the ocean model is accurate and has high enough resolution to resolve ocean features at small enough scales, which is in of itself an active research field.

Ocean model realism (especially at finer scales) correlates with computational power and time.
Thus, until a full-scale ocean model can be run in real-time on a small computer like a Raspberry Pi, the framework proposed in this thesis strikes a balance for the necessary environmental realism as depth-dependent with temporal updates.
This has a cascading effect on acoustic realism, currently a range-independent scheme.
Faster, more effective computational methods (and hardware devices) for acoustic propagation can certainly improve the real-time joint estimation problem, but are still constrained by this depth-dependent snapshot of the ocean volume.
Until we reach that faraway goal, an updatable, range independent scheme, as employed in this work (see Section...) is a practical and effective approach.

The Arctic, a generally inaccessible, dangerous, and expensive region for crewed monitoring via icebreakers,  is a prime location to invest in vehicle deployments and underwater observatories \citep{mikhalevsky_multipurpose_2015}.
As anthropogenic climate change renders the region more accessible, safe navigation infrastructure and practices will be increasingly vital for industry, tourism, commercial, and military operations.

@@

\clearpage
\FloatBarrier
\subsection*{Group velocity prediction by source depth}
\begin{figure}[h!]
\includegraphics[width=0.8\textwidth]{figs/SLIDES-gvel-comparison-wdata-20.pdf} \\
\vspace{2em}
\includegraphics[width=0.8\textwidth]{figs/SLIDES-gvel-comparison-wdata-90.pdf}
\caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing. The top panel shows a source depth of 20 m; the bottom panel shows a source depth of 90 m. The sound speed source is indicated by color. The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}
\label{fig:gvelMore}
\end{figure}
